Speed and strength of an epidemic intervention

### 2020-03-02 doc/speedstrength.tex
very minor tweaks

An epidemic can be characterized by speed (i.e., the exponential growth rate $r$) or strength (i.e., the reproductive number \RR).
Disease modelers have historically placed much more emphasis in strength, in part because the effectiveness of an intervention strategy is typically evaluated on this scale.  
Here we develop a mathematical framework for this classic, strength-based paradigm and show that there is a corresponding speed-based paradigm which can provide complementary insights.
In particular, we note that $r=0$ is a threshold for disease spread, just like $\RR=1$, and show that we can measure the speed and strength of an intervention on the same scale as the speed and strength of an epidemic, respectively.
We argue that, just as the strength-based paradigm provides the clearest insight into certain questions, the speed-based paradigm provides the clearest view in other cases.
As an example, we show that evaluating the prospects of ``test-and-treat'' interventions against the human immunodeficiency virus (HIV) can be done more clearly on the speed than strength scale, given uncertainty in the proportion of HIV spread that happens early in the course of infection. 
We suggest that disease modelers should avoid over-emphasizing the reproductive number at the expense of the exponential growth rate, but instead look at these as complementary measures.

@@

\jd{For next version, consider making \figref{condom} a four-panel figure like \figref{test}.}

@@

Next, we consider a ``test-and-treat'' strategy in which infected individuals are identified, linked to care and receive antiretroviral therapy (ART) with the goal of both preserving health and preventing transmission through viral suppression. \citep{garnett2009treating, granich2009universal, nah2017test}.
Our assumptions for this scenario are shown in \figref{test}.
We assume that the hazard rate $\tsub{h}{test}$ of this intervention starts at 0 (because there is no way for newly infected individuals to know that they have HIV) but increases very quickly (because sexually active individuals are the most likely to seek testing); 
after a few months, the assumed hazard rate goes down to account for the effects of people who avoid identification, persistent treatment failures, and the possibility of rare transmission even under effective treatment \figref{test}(C). The corresponding strength of intervention $L$ is shown in \figref{test}(A) and details of the assumption are given in the caption.

@@

\jd{Let's look at some GIs: high, medium and low \pEarly $\times$ intrinsic, backward}

@@

However, we assume that removal eventually saturates, to account for the effects of people who avoid identification, persistent treatment failures, and the possibility of rare transmission even under effective treatment
(\figref{test}A; see caption for the functional form of $\tsub{L}{test}$).
Under this assumption, the strength of intervention decreases with the proportion of early transmission in a manner roughly similar to the epidemic strength (\figref{test}B).

@@

While both speed- and strength-based frameworks give the same conclusion about the outcome of an intervention, sometimes one provides a clearer understanding of the outcome than the other.
For example, we expect the speed-based framework to be clearer for characterizing newly invading pathogens: when an epidemic is growing exponentially, the reproductive number cannot be estimated with confidence \citep{weitz2015modeling}, especially when there is large uncertainty in the shape of the generation-interval distribution \citep{park2020reconciling},
Conversely, we expect the strength-based framework to be clearer for evaluating established pathogens (based on the effective proportion of the population susceptible).
For interventions, we expect the speed-based framework to be clearer for evaluating intervention strategies that work at the individual level, like test-and-treat for HIV \citep{granich2009universa}, or contact-tracing and quarantine for COVID-19; we expect the strength-based framework to be clearer for intervention strategies that seek to reduce the overall transmission at the population level, like condom use.
In other cases, such as real-time rollout of vaccines during an outbreak, both speed and strength approaches might be similarly uncertain because the result depends both on the speed of the rollout and the (strength-like) final coverage \citep{shah2018mumps}.

>>>>

An epidemic can be characterized by its speed (i.e., the exponential growth rate $r$) and strength (i.e., the reproductive number \RR).
Disease modelers have historically placed much more emphasis in strength, in part because the effectiveness of an intervention strategy is typically evaluated on this scale.  
Here, we develop a mathematical framework for this classic, strength-based paradigm and show that there is a corresponding speed-based paradigm which can provide complementary insights.
In particular, we note that $r=0$ is a threshold for disease spread, just like $\RR=1$, and show that we can measure the speed and strength of an intervention on the same scale as the speed and strength of an epidemic, respectively.
We argue that, just as the strength-based paradigm provides the clearest insight into certain questions, the speed-based paradigm provides the clearest view in other cases.
As an example, we show that evaluating the prospects of ``test-and-treat'' interventions against the human immunodeficiency virus (HIV) can be done more clearly on the speed than strength scale, given uncertainty in the proportion of HIV spread that happens early in the course of infection. 
We suggest that disease modelers should avoid over-emphasizing the reproductive number at the expense of the exponential growth rate, but instead look at these as complementary measures.

@@

% \jd{For next version, consider making \figref{condom} a four-panel figure like \figref{test}.}

@@

Next, we consider a ``test-and-treat'' strategy in which infected individuals are identified, linked to care and receive antiretroviral therapy (ART) with the goal of both preserving health and preventing transmission through viral suppression. \citep{garnett2009treating, granich2009universal, nah2017test}.
Our assumptions for this scenario are shown in \figref{test}.
We assume that the hazard rate $\tsub{h}{test}$ of this intervention starts at 0 (because there is no way for newly infected individuals to know that they have HIV) but increases very quickly (because sexually active individuals are the most likely to seek testing); 
after a few months, the assumed hazard rate goes down to account for the effects of people who avoid identification, persistent treatment failures, and the possibility of rare transmission even under effective treatment (\figref{test}C). The corresponding strength of intervention $L$ is shown in \figref{test}A and details of the assumption are given in the caption.

@@

% \jd{Let's look at some GIs: high, medium and low \pEarly $\times$ intrinsic, backward}

@@

@@

While both speed- and strength-based frameworks give the same conclusion about the outcome of an intervention, sometimes one provides a clearer understanding of the outcome than the other.
For example, we expect the speed-based framework to be clearer for characterizing newly invading pathogens: when an epidemic is growing exponentially, the reproductive number cannot be estimated with confidence \citep{weitz2015modeling}, especially when there is large uncertainty in the shape of the generation-interval distribution \citep{park2020reconciling}.
Conversely, we expect the strength-based framework to be clearer for evaluating established pathogens (based on the effective proportion of the population susceptible).
For interventions, we expect the speed-based framework to be clearer for evaluating intervention strategies that work at the individual level, like test-and-treat for HIV \citep{granich2009universal}, or contact-tracing and quarantine for COVID-19 \citep{hellewell2020feasibility}; we expect the strength-based framework to be clearer for intervention strategies that seek to reduce the overall transmission at the population level, like condom use.
In other cases, such as real-time rollout of vaccines during an outbreak, both speed and strength approaches might be similarly uncertain because the result depends both on the speed of the rollout and the (strength-like) final coverage \citep{shah2018mumps}.
